Transfer learning for echelle spectroscopy

### 2021-11-10 paper/paper1/ms.tex
Significant prose writing today, done for the day/moment

@@

\begin{eqnarray}
  \mathsf{M}(\bm{\lambda}_s) = {\displaystyle \prod_{j=1}^{N_{\mathrm{lines}}}  (1-a_j V(\bm{\lambda}_s-\lambda_{c,j}, \sigma_j, \gamma_j) )}
\end{eqnarray}

@@

The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths,lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.  We minimize a scalar loss function, which we define as the mean squared error (MSE), proportional to $\chi^2$, the sum of the squares of the residual vector, $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$:

@@

Here we employ a variant of Stochastic Gradient Descent (SGD), an optimization technique that can scale to a virtually unlimited number of parameters\authorcomment1{provide a citation}.  This technique computes the derivative of the loss scalar with respect to each of the parameters, the Jacobian: $(\frac{\partial \mathcal{L}}{\partial a_j}, \frac{\partial \mathcal{L}}{\partial \sigma_j}, \frac{\partial \mathcal{L}}{\partial \gamma_j})$.  The Jacobian indicates how the MSE would decrease with a change in the parameter-of-interest, or put simply which way and by how much you have to change the line properties to make a better fit.

@@

In principle we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch.  In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks.  Still, the performance of an analytic implementation over the one presented here represents an interesting avenue of future research.

\subsection{Continuum model}
\begin{outline}
  \1 How to describe the continuum:
  \2 Black body / Rayleigh Jeans plus perturbation
  \2 High order polynomial: Chebyshev? Somewhere between 3rd and 20th order?
  \2 Use cross validation?
  \1 GPs are also possible, but high Number of pixels--> need scalable GP
  \2 Celerite not implemented for PyTorch?
  \2 GPyTorch possible, but has certain restrictions on kernel and sampling
\end{outline}

\subsection{GPU and Autodiff specific considerations}
\begin{outline}
  \1 Total number of parameters
  \2 For the lines =  N-lines x 4 (lambda center, fwhm L, sigma G, Amplitude)
  \2 For the continuum = Nth order polynomial = N+1 parameters, or Gaussian Process = 2 hyperparameters
  \1 Each Tunable parameter gets `grad=True`
  \1 We assemble the product of all lines and all continuum sources, this is the computational bottleneck
  \1 This becomes a matrix of size N-lines x N-wavelength points
  \1 We take the element-wise product over the lines axis, contracting to 1 x N-wavelength points
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*k primative operations.
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*32 bits of memory => XX GB
  \1 NVIDIA GPUs can handle 8 - 40 GB of memory depending on the model
\end{outline}

@@

\subsection{Questioning the assumptions of the method}
\begin{outline}
  \1 Is PseudoVoigt Adequate?
  \1 What performance level is adequate? (depends on your application)
  \1 What about PRV applications
  \1 What about the choice of minibatch size?
  \1 How would a sparse tensor change the computational performance?
  \1 Are we using the GPU effectively?  Are we memory bandwidth limited?
  \1 Should we update the line center positions?
\end{outline}

>>>>

\subsection{The \emph{blas\'e} clone model}

@@

\begin{eqnarray}
  \mathsf{M}(\bm{\lambda}_s) = {\displaystyle \prod_{j=1}^{N_{\mathrm{lines}}}  (1-a_j V(\bm{\lambda}_s-\lambda_{c,j}, \sigma_j, \gamma_j) )} \label{equation1}
\end{eqnarray}

@@

The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.  We minimize a scalar ``goodness-of-fit'' metric, \emph{aka}loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$ but has no notion of per-pixel noise, since the precomputed synthetic spectrum has no uncertainty:

@@

Here we employ a variant of Stochastic Gradient Descent (SGD), an optimization technique that can scale to a virtually unlimited number of parameters\authorcomment1{provide a citation}.  This technique computes the derivative of the loss scalar with respect to each of the parameters, the Jacobian: $(\frac{\partial \mathcal{L}}{\partial a_j}, \frac{\partial \mathcal{L}}{\partial \sigma_j}, \frac{\partial \mathcal{L}}{\partial \gamma_j})$.  The Jacobian indicates how the MSE would decrease with a change in the parameter-of-interest, or put simply which-way and by-how-much you have to change the line properties to get a better fit.

@@

In principle we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch.  Futher, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting.  In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks.  The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation.  Still, the performance of an analytic Jacobian implementation over the autodiff route presented here presents an interesting avenue of future research.

\subsection{GPU and Autodiff specific considerations}
We make a few tweaks to the implementation for numerical purposes.  First, we want all the parameters to be positive, forbidding negative amplitudes and negative widths.  We therefore tune the natural log of the parameters, and exponentiate them before inclusion in Equation \ref{equation1}.  Second, we found through iterative experimentation that the initialization amplitudes and widths were systematically shifted from the optimized values.  We built-in a three scalar tweaks to the initialization amplitudes, Gaussian width, and Lorentzian width, which dramatically accelerated the optimization step.

We set the \texttt{requires\_grad=True} property for any Torch tensor that we want to vary.  This allows us to easily explore whether, say, allowing the $\lambda_c$ parameter to vary significantly improves the fit.  It also allows us to tune or freeze an experimental wavelength-dependent pre-factor to Equation \ref{equation1} to correct for lingering imperfections in the otherwise-fixed continuum flattening procedure.

The computational bottleneck occurs at...

\begin{outline}
  \1 Total number of parameters
  \2 For the lines =  N-lines x 4 (lambda center, fwhm L, sigma G, Amplitude)
  \2 For the continuum = Nth order polynomial = N+1 parameters, or Gaussian Process = 2 hyperparameters
  \1 We assemble the product of all lines and all continuum sources, this is the computational bottleneck
  \1 This becomes a matrix of size N-lines x N-wavelength points
  \1 We take the element-wise product over the lines axis, contracting to 1 x N-wavelength points
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*k primative operations.
  \1 For N-lines = X and N-wavelength-points = Y, we expect X*Y*32 bits of memory => XX GB
  \1 NVIDIA GPUs can handle 8 - 40 GB of memory depending on the model
\end{outline}

@@

\subsection{Questioning the assumptions of the method}

\begin{outline}
  \1 Is PseudoVoigt Adequate?
  \1 What performance level is adequate? (depends on your application)
  \1 What about PRV applications
  \1 What about the choice of minibatch size?
  \1 How would a sparse tensor change the computational performance?
  \1 Are we using the GPU effectively?  Are we memory bandwidth limited?
  \1 Should we update the line center positions?
\end{outline}
